Identifying Microlensing Events in Large, Non-Uniformly Sampled Surveys: The Case of the Palomar Transient Factory

### 2013-01-24 text/paper/ptf_microlensing.tex

%\documentclass{emulateapj}
\documentclass[12pt,preprint]{aastex}
\newcounter{address}
\setcounter{address}{1}
%\usepackage{lscape, longtable}
\usepackage{lmodern}
%\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{subfigure}
\usepackage{natbib}

@@

\section{PTF } % Better section header?
The Palomar Transient Factory is a transient detection system incorporating three elements. The first is a wide-field survey camera mounted on the automated 48 inch Oschin Schmidt telescope (the P48) at Palomar Observatory, CA. This is the former Canada-France-Hawaii Telescope 12K$\times$8K mosaic camera, which has 11 working chips, 10$^7$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}. Under median seeing conditions (1.1$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey \citep[SDSS;][]{york00} $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in a standard 60~s exposure \citep{nick2010}. As of September 2012, the PTF footprint included $\apwsim$10,100 (2300) deg$^2$ imaged $>$25 ($>$100) times in $R$ and $\apwsim$2200 (170) deg$^2$ imaged that often in $g$ (\figurename~\ref{fig:survey_footprint})\footnote{For an interactive visualization of the PTF survey coverage, see: \url{http://bit.ly/PTFSky}}. PTF is also a real-time data reduction pipeline that identifies transients of interest, passes these to a dedicated photometric follow-up telescope, and generates an archive of all detected sources \citep{nick2009,rau2009}. %[CHECK NUMBERS]

%  [WE ALSO PROBABLY WANT TO CITE A FEW OF THE ACTUAL RESULT PAPERS; WE CAN WORRY ABOUT THAT LATER]
The PTF survey footprint is not uniformly sampled either spatially or temporally. Each field has a unique sampling pattern determined by (1) which of the PTF sub-surveys it belongs to, (2) which time of the year it is visible from Palomar, and (3) how high a priority it is given by the scheduler. The survey design strategy was to allocate 92\% of the P48 time between three surveys: a 5-day cadence experiment covering 8000~deg$^2$ at $|b|>30^\circ$ to search for extragalactic transients, a dynamic cadence experiment to identify  short duration ($<$5~days) transients, and a high-cadence imaging campaign of a single field in Orion to find transiting Jupiter-mass planets around young stars \citep{nick2009}. As a result, PTF light curves often contain gaps, regions of high-cadence observations, and regions of low-cadence observations, leading to a massive data set of irregularly-sampled, time-domain photometry. \figurename~\ref{fig:sampling} shows six randomly-selected light curves and demonstrates the varying cadences and coverage that different fields may have over the same one-year baseline. We restrict our sample to the \apwsim$6.4\times10^7$ light curves with $>$25 \textit{R}-band observations.

@@

The parameters for each simulated event are chosen as follows: $t_0$ is drawn from a uniform distribution between the minimum and maximum observation times, $u_0$ is drawn from a uniform distribution between 0 and 1.34 ([TODO:] cite?), and $t_E$ is drawn from a log-uniform distribution between 1 and 1000 days. For each iteration we recompute the variability statistics and store the event parameters. Figures~\ref{fig:detection_efficiency_4327}-\ref{fig:detection_efficiency_100152} show detection efficiency curves (bottom panels) computed using this simulation for three representative fields chosen for their different sampling patterns (top panels). We ignore the indices \textit{K} and $\sigma/\mu$ because their integrated detection efficiencies are each below 1\%. For each of these fields we find that $\eta$ consistently performs better than the other indices at recovering microlensing events. This is especially interesting because computing $\eta$ is \apwsim100 times faster per light curve as compared to $\Delta\chi^2$, and that identifying candidate microlensing events in the full PTF database using this statistic is therefore computationally plausible.

@@

\section{Searching for events in the full PTF data set}
% FLUSH OUT AND UPDATE THIS SECTION
As an initial test, we applied the 1\% FPR selection on the 325 PTF fields (\apwsim2400 deg$^2$) with $>$100 $R$-band observations using the variability index $\eta$. Our procedure is as follows:
\begin{enumerate}
	\item select a field from the sample of fields with $>100$ good $R$-band observations, and run Monte Carlo simulations to compute the 1\% FPR limiting values for the variability indices;
	\item compute $\eta$ for each light curve, and select all light curves that pass the cut determined from the previous simulations;
	\item fit a microlensing event model and subtract the model, then recompute $\eta$ for each selected light curve;
	\item we reject the light curve if the new value of $\eta$ still passes the cut (it is probably a periodic variable);
	\item any remaining light curves are loaded into a local database to be vetted by hand.
\end{enumerate}
% [THE ABOVE WOULD MAKE A VERY NICE FLOWCHART] --> for the actual paper..

>>>>

\documentclass{emulateapj}
%\documentclass[12pt,preprint]{aastex}
\newcounter{address}
\setcounter{address}{1}
%\usepackage{lscape, longtable}
\usepackage{lmodern}
%\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{subfigure}
\usepackage{natbib}

@@

\section{PTF } % Better section header?
The Palomar Transient Factory is a transient detection system incorporating three elements. The first is a wide-field survey camera mounted on the automated 48 inch Oschin Schmidt telescope (the P48) at Palomar Observatory, CA. This is the former Canada-France-Hawaii Telescope 12K$\times$8K mosaic camera, which has 11 working chips, 10$^7$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}. Under median seeing conditions (1.1$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey \citep[SDSS;][]{york00} $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in a standard 60~s exposure \citep{nick2010}. As of November 2012, the PTF footprint included $\apwsim$15,224 (2766) deg$^2$ imaged $>$10 ($>$100) times in $R$ and $\apwsim$5430 (290) deg$^2$ imaged that often in $g$ (\figurename~\ref{fig:survey_footprint})\footnote{For an interactive visualization of the PTF survey coverage, see: \url{http://bit.ly/PTFSky}}. PTF is also a real-time data reduction pipeline that identifies transients of interest, passes these to a dedicated photometric follow-up telescope, and generates an archive of all detected sources \citep{nick2009,rau2009}.

%  [WE ALSO PROBABLY WANT TO CITE A FEW OF THE ACTUAL RESULT PAPERS; WE CAN WORRY ABOUT THAT LATER]
The PTF survey footprint is not uniformly sampled either spatially or temporally. Each field has a unique sampling pattern determined by (1) which of the PTF sub-surveys it belongs to, (2) which time of the year it is visible from Palomar, and (3) how high a priority it is given by the scheduler. The survey design strategy was to allocate 92\% of the P48 time between three surveys: a 5-day cadence experiment covering 8000~deg$^2$ at $|b|>30^\circ$ to search for extragalactic transients, a dynamic cadence experiment to identify  short duration ($<$5~days) transients, and a high-cadence imaging campaign of a single field in Orion to find transiting Jupiter-mass planets around young stars \citep{nick2009}. As a result, PTF light curves often contain gaps, regions of high-cadence observations, and regions of low-cadence observations, leading to a massive data set of irregularly-sampled, time-domain photometry. \figurename~\ref{fig:sampling} shows six randomly-selected light curves and demonstrates the varying cadences and coverage that different fields may have over the same one-year baseline. [APW:] Figure ?? shows distributions of a) total number of $R$-band observations and b) field baseline for all fields in PTF. We restrict our sample to the \apwsim$1.1\times10^9$ light curves with $>$10 \textit{R}-band observations because there is a drop off in the number of fields with fewer than 10 observations and we don't expect such light curves to be useful for our search. %APW!

@@

The parameters for each simulated event are chosen as follows: $t_0$ is drawn from a uniform distribution between the minimum and maximum observation times, $u_0$ is drawn from a uniform distribution between 0 and 1.34 ([TODO:] cite?), and $t_E$ is drawn from a log-uniform distribution between 1 and 1000 days. For each iteration we recompute the variability statistics and store the event parameters. Figures~\ref{fig:detection_efficiency_4327}-\ref{fig:detection_efficiency_100152} show the results for three representative fields chosen for their different sampling patterns (top panels) and detection efficiency curves computed using this simulation (bottom panels). We ignore the indices \textit{K} and $\sigma/\mu$ because their integrated detection efficiencies are each below 1\%. For each of these fields we find that $\eta$ consistently performs better than the other indices at recovering microlensing events. This is especially interesting because computing $\eta$ is \apwsim100 times faster per light curve as compared to $\Delta\chi^2$, and that identifying candidate microlensing events in the full PTF database using this statistic is therefore computationally plausible.

@@

\section{Searching for events in the full PTF data set}
[APW:  FLUSH OUT AND UPDATE THIS SECTION when pipeline tests are done]

As an initial test, we applied the 1\% FPR selection on the 325 PTF fields (\apwsim2400 deg$^2$) with $>$100 $R$-band observations using the variability index $\eta$. Our procedure is as follows:
\begin{enumerate}
	\item select a field from the sample of fields with $>100$ good $R$-band observations, and run Monte Carlo simulations to compute the 1\% FPR limiting values for the variability indices;
	\item compute $\eta$ for each light curve, and select all light curves that pass the cut determined from the previous simulations;
	\item fit a microlensing event model and subtract the model, then recompute $\eta$ for each selected light curve;
	\item we reject the light curve if the new value of $\eta$ still passes the cut (it is probably a periodic variable);
	\item any remaining light curves are loaded into a local database to be vetted by hand.
\end{enumerate}
% [THE ABOVE WOULD MAKE A VERY NICE FLOWCHART] --> for the actual paper..
